Energy Clustering

### 2017-09-17 qcqp_clustering/draft/energy_clustering.tex
finished round of comments

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_dim.pdf}\\[-1.0em] (a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_cov_squareroot.pdf}\\[-1.0em] (b)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_cov_linear.pdf}\\[-1.0em] (c)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_cov_square.pdf}\\[-1.0em] (d)
\end{minipage}
\caption{
\label{fig:gauss}
Comparison of Algorithm~\ref{algo}, Algorithm~\ref{kmeans_algo},
standard $k$-means and GMM/EM as the number of dimensions increase
in gaussian settings.
We compute the average of \eqref{eq:accuracy} over $100$ samples with
error bars being standard error. We have two clusters with $100$ points
each. 
(a) Data as in \eqref{eq:gauss1}, where the optimal accuracy
from Bayes error is the dashed line equal to $\approx 0.86$.
(b) Data from \eqref{eq:gauss2} with $q=1/2$ in \eqref{eq:cov}. 
(c) Data from \eqref{eq:gauss2} with $q=1$ in \eqref{eq:cov}.
(d) Data from \eqref{eq:gauss2} with $q=2$ in \eqref{eq:cov}.
The optimal accuracy from Bayes error in (b--d) is $\approx 1$.
% @gui: titles for each, or subcaptions, would be extremely useful.  in general, titles should provide the "setting" of the experiment, ie, the model, etc.
% @gui: also helpful here would be pairs plots, say, of the first 5 dimensions, showing the performance of our stuff (color points by Alg 3, maybe symbols denote "truth".  i have no intuition how difficult these are.
}
\end{figure}

@@

To summarize, in the experiments of Fig.~\ref{fig:gauss}
we see a better performance of Algorithm~\ref{algo} compared
to the other ones, and in particular to kernel $k$-means,
where we recall that both find local solutions
to the same optimization problem \eqref{eq:qcqp2}.
Algorithm~\ref{algo} is more robust as the number of dimensions
increase.

Let us now
consider the effect of having 
unbalanced clusters. We generate data as
\begin{equation}
\label{eq:gauss3}
\begin{split}
x &\in \C_i \sim  
\dfrac{n_i}{N} \mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2), \quad 
\mu_1 = (0,0,0,0)^\top , \quad
\mu_2 = 1.5\times (1,1,0,0)^\top, \\
\Sigma_1 &= I_4, \quad
\Sigma_2 = \left( 
\begin{smallmatrix} 
1/2 & 0 & 0 & 0\\
0 & 1/2 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 
\end{smallmatrix}\right), \quad
n_1 = N - m, \quad  n_2 = N + m, \quad N=200.
\end{split}
\end{equation}
We then increase $m$, 
that is we make the clusters progressively more unbalanced,
and plot the average of \eqref{eq:accuracy} 
over
$100$ samples for each $m$
(error bars are standard error). The results are in Fig.~\ref{fig:unbalanced}.
As expected, GMM
works better than the other algorithms since it is a fuzzy clustering.
The other methods based on hard assignments degrade similarly
and  more rapidly than GMM. This indicates that a fuzzy version of
Algorithm~\ref{algo} should compensate for this effect.

@@

Besides the standard kernel from energy statistics 
\eqref{eq:standard_metric}, 
let us consider two other semimetrics with their respective generating
kernels,
\begin{align}
\rho_{1/2}(x,y) &= \| x-y \|^{1/2} & 
 \kk_{1/2}(x,y) &= \tfrac{1}{2} \left( 
\| x \|^{1/2} + \| y \|^{1/2} 
- \| x-y \|^{1/2} \right), \label{eq:rhohalf}\\
\rho_{e}(x,y) &= 
2 - 2 e^{-\tfrac{1}{2}\| x- y\|} &
 \kk_{e}(x,y) &= e^{-\tfrac{1}{2}\| x-y\|}.
\label{eq:rhoe}
\end{align}
The kernel \eqref{eq:rhohalf} corresponds to the energy distance
\eqref{eq:energy2} with $\alpha=1/2$.
Now consider data normally distributed in $D=20$ as
\begin{equation}
\label{eq:20gauss}
\begin{split}
x &\in \C_i \sim \mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2), \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top ,\quad
\mu_2 = \tfrac{1}{2} 
(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top, \quad
\Sigma_1 = \tfrac{1}{2} I_{20},  \quad
\Sigma_2 = I_{20}.
\end{split}
\end{equation}
We sample an equal number of points for each cluster, which is progressively
increased. The optimal accuracy based on Bayes
classification error is $\approx 0.90$. 
Clustering results are shown in Fig.~\ref{fig:consist}a.
Algorithm~\ref{algo} outperforms the other ones, and in 
particular the kernel \eqref{eq:rhoe} provides better results.
As the number of points get large enough GMM starts to approach
optimal Bayes, as it should since it is a
consistent model to the data. However, 
Algorithm~\ref{algo} with kernel \eqref{eq:rhoe} approach optimal Bayes
with a much smaller number of points. Moreover, Algorithm~\ref{algo} 
outperforms
Algorithm~\ref{kmeans_algo} for any of the kernel choices.
%Standard $k$-means performs poorly in this example.

Now consider the same parameters as in \eqref{eq:20gauss} but with
lognormal distributions, 
\begin{equation}
\label{eq:20loggauss}
\log x \in \C_i \sim \mathcal{N}(\mu_i, \Sigma_i) \qquad (i=1,2).
\end{equation}
The same previous experiment is shown in 
Fig.~\ref{fig:consist}b.
Note that Algorithm~\ref{algo} still performs accurately, 
while $k$-means works almost at chance,
and GMM is not even able to estimate the parameters. Again, the
kernel \eqref{eq:rhoe}
provides better results than \eqref{eq:standard_metric} or
\eqref{eq:rhohalf}. 
The experiments in Fig.~\ref{fig:consist}
illustrate how energy statistics clustering is nonparametric.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{gauss.pdf}\\[-1.0em]
(a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{loggauss.pdf}\\[-1.0em]
(b)
\end{minipage}
\caption{
\label{fig:consist}
Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} with kernels
\eqref{eq:standard_metric}, \eqref{eq:rhohalf} and \eqref{eq:rhoe}, 
$k$-means, and GMM. The optimal accuracy in both cases
is $\approx 0.9$. We show the average of \eqref{eq:accuracy}
over $100$ samples with standard error.
(a) Data distributed as in 
\eqref{eq:20gauss}. 
(b) Data distributed as in \eqref{eq:20loggauss}.
}
\end{figure}

@@

Consider the following choices of semimetric and 
corresponding generating kernel:
\begin{align}
\rho_{\alpha}(x,y) &= \| x - y\|^\alpha & 
\kk_{\alpha}(x,y) &= \tfrac{1}{2}\left(
\| x \|^\alpha +
\| y \|^\alpha -
\| x-y \|^\alpha \right) ,
\label{eq:rhoalpha} \\
%
\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}} &
\widetilde{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|}{2\sigma}} ,
\label{eq:rhoesigma} \\
%
\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}} &
\widehat{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|^2}{2\sigma^2}} .
\label{eq:rhogsigma} 
\end{align}
In Fig.~\ref{fig:other} we have examples of
complex two dimensional datasets. The two parallel cigars of 
Fig.~\ref{fig:other}a have $200$ points each. For the concentric circles
of Fig.~\ref{fig:other}b and Fig.~\ref{fig:other}c we
sample $400$ points for each class.
We apply Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo}, besides
$k$-means and GMM, using the above kernels 
\eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}. The results are shown
in Table~\ref{table:other} where the respective choice of parameters for the
kernels are indicated. For the data in Fig.~\ref{fig:other}a 
the semimetrics $\rho_1$ and $\rho_{1/2}$ are
able to provide more accurate results compared to $k$-means. However,
the gaussian kernel $\widetilde{\rho}_2$ gives very accurate
results, similar to GMM,
which is the true model of the data. For the data shown in
Fig.~\ref{fig:other}b we see
that the choice of kernel is very sensitive, and only \eqref{eq:rhoesigma}
was able to cluster accurately. The same kernel choice to the case
of Fig.~\ref{fig:other} still provides better results than
the others, but they are less accurate compared to Fig.~\ref{fig:other}b.

\newcolumntype{g}{>{\columncolor{gray!20}}l}
\begin{table}[h]
\renewcommand*{\arraystretch}{0.75}
\begin{tabular}{@{}r  l g  l g  l g@{}}
\toprule[1pt]
 & & \emph{Fig.~\ref{fig:other}a}
 & & \emph{Fig.~\ref{fig:other}b}
 & & \emph{Fig.~\ref{fig:other}c} \\
\midrule[0.5pt]
\multirow{4}{*}{\emph{Algorithm \ref{algo}}~~~~} 
& $\rho_{1}$ & $0.766\pm0.066$
& $\rho_{1}$ & $0.522\pm0.006$
& $\rho_{1}$ & $0.437\pm0.030$ \\
& $\rho_{1/2}$ & $0.859\pm0.062$
& $\rho_{1/2}$ & $0.524\pm0.007$
& $\rho_{1/2}$ & $0.547\pm0.026$ \\
& $\widetilde{\rho}_{2}$ & $0.971\pm0.015$
& $\widetilde{\rho}_{1}$ & $\bm{0.9999\pm0.0001}$
& $\widetilde{\rho}_{2}$ & $\bm{0.677\pm0.003}$ \\
& $\widehat{\rho}_{2}$  & $\bm{0.998\pm0.001}$
& $\widehat{\rho}_{1}$  & $0.597\pm0.052$
& $\widehat{\rho}_{2}$ & $0.645\pm0.012$ \\
\arrayrulecolor{gray!80}\midrule[0.5pt]
\multirow{4}{*}{\emph{Algorithm \ref{kmeans_algo}}~~~~} 
& $\rho_{1}$ & $0.758\pm 0.069$
& $\rho_{1}$ & $0.516\pm 0.002$
& $\rho_{1}$  & $0.452\pm 0.030$ \\
& $\rho_{1/2}$ & $0.901\pm 0.060$
& $\rho_{1/2}$ & $0.524\pm 0.007$
& $\rho_{1/2}$ & $0.570\pm 0.016$ \\
& $\widetilde{\rho}_{2}$ & $0.971\pm 0.015$
& $\widetilde{\rho}_{1}$ & $\bm{0.9999\pm 0.0001}$
& $\widetilde{\rho}_{2}$ & $\bm{0.673\pm 0.002}$ \\
& $\widehat{\rho}_{2}$ & $\bm{0.998\pm 0.001}$
& $\widehat{\rho}_{1}$ & $0.528\pm 0.008$
& $\widehat{\rho}_{2}$ & $0.640\pm 0.013$ \\
\arrayrulecolor{gray!80}\midrule[0.5pt]
\emph{$k$-means}~~~~ 
& & $0.599\pm0.046$
& & $0.521\pm0.005$
& & $0.360\pm0.004$ \\
\emph{GMM}~~~~
& & $0.9995\pm0.0003$
& & $0.598\pm0.018$
& & $0.479\pm0.021$ \\
\arrayrulecolor{black}\bottomrule[1pt]
\end{tabular}
\caption{\label{table:other}
Clustering the data shown in Fig.~\ref{fig:other} with Algorithm~\ref{algo}
and Algorithm~\ref{kmeans_algo}, besides $k$-means and GMM, with kernels 
\eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}. 
We sample $10$ times and show the average
accuracy \eqref{eq:accuracy} with standard error.
}
\end{table}

Next we consider 
the well-known MNIST handwritten digit dataset 
as illustrated in Fig.~\ref{fig:other}d.
Each data point is an $8$-bit gray scale
image forming a $784$-dimensional vector 
corresponding to the digits $\{0,1,\dotsc,9 \}$.
Besides the kernel \eqref{eq:rhoalpha}, we
consider the gaussian kernel \eqref{eq:rhogsigma} with 
\begin{equation}
\label{eq:sigma}
\sigma^2 = \dfrac{1}{n^2} \sum_{i,j=1}^n \| x_i - x_j \|^2 ,
\end{equation}
which is computed from a sample  $\{ x_i \}_{i=1}^n$.
We consider subsets of the  classes $\{0,1,\dotsc,9 \}$, 
where we sample $100$ points 
for each class. We perform clustering through 
Algorithm~\ref{algo},
Algorithm~\ref{kmeans_algo}, and $k$-means.
The results are shown in Table~\ref{table:mnist} where the kernel
and its parameter for each case is indicated.
Algorithm~\ref{algo} performed slightly better than $k$-means, except
for the last column where all the methods are comparable. 
Unsupervised clustering on MNIST dataset without any feature extraction
or dimensionality reduction is not an easy task. For instance,
the same experiment was performed in \cite{Sapiro} where a low-rank
transformation is learned then subsequently used in subspace clustering,
providing very accurate results. One could explore analogous methods
for learning a better representation of the data and subsequently apply
Algorithm~\ref{algo} for clustering.

\newcolumntype{g}{>{\columncolor{gray!20}}l}
\begin{table}[h]
\renewcommand*{\arraystretch}{0.75}
\begin{tabular}{@{}rl|l|l|l|l@{}}
\toprule[1pt]
\emph{Class Subset}~~~~ & & $\{0,1,2\}$ &
$\{0,1,\dotsc,4\}$ &
$\{0,1,\dotsc,6\}$ &
$\{0,1,\dotsc,8\}$ \\
\midrule[0.5pt]
\multirow{3}{*}{\emph{Algorithm \ref{algo}}~~~~} 
& $\rho_{1}$\hspace{1em} 
&$0.907\pm 0.007$
&$0.866\pm 0.006$
&$0.715\pm 0.013$
&$0.616\pm 0.019$
\\
& $\rho_{1/2}$ 
&$\bm{0.918\pm 0.006}$
&$0.849\pm 0.025$
&$0.711\pm 0.010$
&$0.642\pm 0.009$
\\
& $\widetilde{\rho}_{\sigma}$ 
&$0.900\pm 0.007$
&$\bm{0.871\pm 0.005}$
&$\bm{0.719\pm 0.010}$
&$0.630\pm 0.016$
\\
\arrayrulecolor{gray!80}\midrule[0.5pt]
\multirow{3}{*}{\emph{Algorithm \ref{kmeans_algo}}~~~~} 
& $\rho_{1}$ 
&$0.914\pm 0.006$
&$0.845\pm 0.023$
&$0.664\pm 0.022$
&$0.614\pm 0.014$
\\
& $\rho_{1/2}$ 
&$0.895\pm 0.011$
&$0.822 \pm 0.026$
&$0.669\pm 0.021$
&$0.591\pm 0.019$
\\
& $\widetilde{\rho}_{\sigma}$ 
&$0.896\pm 0.007$
&$0.869\pm 0.006$
&$0.705\pm 0.016$
&$\bm{0.646\pm 0.020}$
\\
\arrayrulecolor{gray!80}\midrule[0.5pt]
\emph{$k$-means}~~~~ &
&$0.871\pm 0.015$
&$0.840\pm 0.022$
&$0.707\pm 0.012$
&$0.634\pm 0.011$
\\
\arrayrulecolor{black}\bottomrule[1pt]
\end{tabular}
\caption{\label{table:mnist}
Clustering the data shown in Fig.~\ref{fig:other}d with Algorithm~\ref{algo},
Algorithm~\ref{kmeans_algo}, and $k$-means.
We use the kernel \eqref{eq:rhoalpha} with $\alpha\in\{1,2\}$ and
the gaussian kernel 
\eqref{eq:rhogsigma} with $\sigma$ given by \eqref{eq:sigma}. 
For each subset of digits we sample $10$ times and show the average
accuracy \eqref{eq:accuracy} with standard error. We sample $100$ points for 
each class.
}
\end{table}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
\label{sec:conclusion}

>>>>

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_dim.pdf}\\[-1.0em] (a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_cov_squareroot.pdf}\\[-1.0em] (b)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_cov_linear.pdf}\\[-1.0em] (c)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_cov_square.pdf}\\[-1.0em] (d)
\end{minipage}
\caption{
\label{fig:gauss}
Comparison of Algorithm~\ref{algo}, Algorithm~\ref{kmeans_algo},
standard $k$-means and GMM/EM as the number of dimensions increase
in gaussian settings.
% @gui: gaussian is always captilized
% @gui: also, i don't understand this plot entirely, are you really starting with D=1? or 10? from the equations, it would seem that Bayes error changes when D< 10, but your Bayes error lines do not seem to change?
We compute the average of \eqref{eq:accuracy} over $100$ samples with
error bars being standard error. We have two clusters with $100$ points
each. 
(a) Data as in \eqref{eq:gauss1}, where the optimal accuracy
from Bayes error is the dashed line equal to $\approx 0.86$.
(b) Data from \eqref{eq:gauss2} with $q=1/2$ in \eqref{eq:cov}. 
(c) Data from \eqref{eq:gauss2} with $q=1$ in \eqref{eq:cov}.
(d) Data from \eqref{eq:gauss2} with $q=2$ in \eqref{eq:cov}.
The optimal accuracy from Bayes error in (b--d) is $\approx 1$.
% @gui: titles for each, or subcaptions, would be extremely useful.  in general, titles should provide the "setting" of the experiment, ie, the model, etc.
% @gui: also helpful here would be pairs plots, say, of the first 5 dimensions, showing the performance of our stuff (color points by Alg 3, maybe symbols denote "truth".  i have no intuition how difficult these are.
}
\end{figure}

@@

To summarize, in the experiments of Fig.~\ref{fig:gauss}
we see a better performance of Algorithm~\ref{algo} compared
to the other ones, and in particular to kernel $k$-means,
where we recall that both find local solutions
to the same optimization problem \eqref{eq:qcqp2}.
Algorithm~\ref{algo} is more robust as the number of dimensions
increase.
% @gui: is there any reason to initialize alg 3 with alg 2? is it even worth trying?

Let us now
consider the effect of having 
unbalanced clusters. We generate data as
\begin{equation}
\label{eq:gauss3}
\begin{split}
x &\in \C_i \sim  
\dfrac{n_i}{N} \mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2), \quad 
\mu_1 = (0,0,0,0)^\top , \quad
\mu_2 = 1.5\times (1,1,0,0)^\top, \\
\Sigma_1 &= I_4, \quad
\Sigma_2 = \left( 
\begin{smallmatrix} 
1/2 & 0 & 0 & 0\\
0 & 1/2 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 
\end{smallmatrix}\right), \quad
n_1 = N - m, \quad  n_2 = N + m, \quad N=200.
\end{split}
\end{equation}
We then increase $m$, 
that is we make the clusters progressively more unbalanced,
and plot the average of \eqref{eq:accuracy} 
over
$100$ samples for each $m$
(error bars are standard error). The results are in Fig.~\ref{fig:unbalanced}.
As expected, GMM
works better than the other algorithms since it is a fuzzy clustering.
The other methods based on hard assignments degrade similarly
and  more rapidly than GMM. This indicates that a fuzzy version of
% @gui: in stats, we don't say "fuzzy", we say "soft".  also, why do you think that when clusters are imbalanced that soft clustering is more important? i would say that soft clustering is more important when the cluster centers are close, i don't see how imbalanced matters.
Algorithm~\ref{algo} should compensate for this effect.

@@

Besides the standard kernel from energy statistics 
\eqref{eq:standard_metric}, 
% @gui: please remove all "let us" from manuscript
let us consider two other semimetrics with their respective generating
kernels,
\begin{align}
\rho_{1/2}(x,y) &= \| x-y \|^{1/2} & 
 \kk_{1/2}(x,y) &= \tfrac{1}{2} \left( 
\| x \|^{1/2} + \| y \|^{1/2} 
- \| x-y \|^{1/2} \right), \label{eq:rhohalf}\\
\rho_{e}(x,y) &= 
2 - 2 e^{-\tfrac{1}{2}\| x- y\|} &
 \kk_{e}(x,y) &= e^{-\tfrac{1}{2}\| x-y\|}.
\label{eq:rhoe}
\end{align}
The kernel \eqref{eq:rhohalf} corresponds to the energy distance
\eqref{eq:energy2} with $\alpha=1/2$.
Now consider data normally distributed in $D=20$ as
\begin{equation}
\label{eq:20gauss}
\begin{split}
x &\in \C_i \sim \mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2), \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top ,\quad
\mu_2 = \tfrac{1}{2} 
(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top, \quad
\Sigma_1 = \tfrac{1}{2} I_{20},  \quad
\Sigma_2 = I_{20}.
\end{split}
\end{equation}
We sample an equal number of points for each cluster, which is progressively
increased. The optimal accuracy based on Bayes
classification error is $\approx 0.90$. 
Clustering results are shown in Fig.~\ref{fig:consist}a.
Algorithm~\ref{algo} outperforms the other ones, and in 
particular the kernel \eqref{eq:rhoe} provides better results.
As the number of points get large enough GMM starts to approach
optimal Bayes, as it should since it is a
consistent model to the data. However, 
Algorithm~\ref{algo} with kernel \eqref{eq:rhoe} approach optimal Bayes
with a much smaller number of points. Moreover, Algorithm~\ref{algo} 
outperforms
Algorithm~\ref{kmeans_algo} for any of the kernel choices.
% @gui: this is a separate point, that i would make a separate figure for if you want to make it. then, one fig would show E-H with 3 kernels vs. kmeans & gmm, and then another would show the difference between accuracy of E-H vs E-L for each of the 3 kernels.

Now consider the same parameters as in \eqref{eq:20gauss} but with
lognormal distributions, 
\begin{equation}
\label{eq:20loggauss}
\log x \in \C_i \sim \mathcal{N}(\mu_i, \Sigma_i) \qquad (i=1,2).
\end{equation}
The same previous experiment is shown in 
Fig.~\ref{fig:consist}b.
Note that Algorithm~\ref{algo} still performs accurately, 
while $k$-means works almost at chance,
and GMM is not even able to estimate the parameters. 
% @gui: something is weird with the way you care computing accuracy.  GMM should not be significantly worse than chance.  i think you are probably counting the NaN's as *wrong*, rather than nothing?
Again, the
kernel \eqref{eq:rhoe}
provides better results than \eqref{eq:standard_metric} or
\eqref{eq:rhohalf}. 
The experiments in Fig.~\ref{fig:consist}
illustrate how energy statistics clustering is nonparametric.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{gauss.pdf}\\[-1.0em]
(a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{loggauss.pdf}\\[-1.0em]
(b)
\end{minipage}
\caption{
\label{fig:consist}
% @gui: too many lines, you demonstrated that E-H is better than E-L already, i don't think we need to show the E-L variants here. 
Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} with kernels
\eqref{eq:standard_metric}, \eqref{eq:rhohalf} and \eqref{eq:rhoe}, 
$k$-means, and GMM. The optimal accuracy in both cases
is $\approx 0.9$. We show the average of \eqref{eq:accuracy}
over $100$ samples with standard error.
(a) Data distributed as in 
\eqref{eq:20gauss}. 
(b) Data distributed as in \eqref{eq:20loggauss}.
}
\end{figure}

@@

Consider the following choices of semimetric and 
corresponding generating kernel:
\begin{align}
\rho_{\alpha}(x,y) &= \| x - y\|^\alpha & 
\kk_{\alpha}(x,y) &= \tfrac{1}{2}\left(
\| x \|^\alpha +
\| y \|^\alpha -
\| x-y \|^\alpha \right) ,
\label{eq:rhoalpha} \\
%
\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}} &
\widetilde{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|}{2\sigma}} ,
\label{eq:rhoesigma} \\
%
\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}} &
\widehat{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|^2}{2\sigma^2}} .
\label{eq:rhogsigma} 
\end{align}
In Fig.~\ref{fig:other} we have examples of
complex two dimensional datasets. The two parallel cigars of 
Fig.~\ref{fig:other}a have $200$ points each. For the concentric circles
of Fig.~\ref{fig:other}b and Fig.~\ref{fig:other}c we
sample $400$ points for each class.
We apply Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo}, using the above kernels 
\eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}, as well as
$k$-means and GMM. The results are shown
in Table~\ref{table:other} where the respective choice of parameters for the
kernels are indicated. For the data in Fig.~\ref{fig:other}a 
the semimetrics $\rho_1$ and $\rho_{1/2}$ are
able to provide more accurate results compared to $k$-means. However,
the gaussian kernel $\widetilde{\rho}_2$ gives very accurate
results, similar to GMM,
which is the true model of the data. For the data shown in
Fig.~\ref{fig:other}b we see
that the choice of kernel is very sensitive, 
% @gui: you mean performance is sensitive to the kernel?
and only \eqref{eq:rhoesigma}
was able to cluster accurately. The same kernel choice to the case
of Fig.~\ref{fig:other} 
% @gui: Fig 5 a? 5c?
still provides better results than
the others, but they are less accurate compared to Fig.~\ref{fig:other}b.

\newcolumntype{g}{>{\columncolor{gray!20}}l}
\begin{table}[h]
\renewcommand*{\arraystretch}{0.75}
\begin{tabular}{@{}r  l g  l g  l g@{}}
\toprule[1pt]
 & & \emph{Fig.~\ref{fig:other}a}
 & & \emph{Fig.~\ref{fig:other}b}
 & & \emph{Fig.~\ref{fig:other}c} \\
\midrule[0.5pt]
\multirow{4}{*}{\emph{Algorithm \ref{algo}}~~~~} 
& $\rho_{1}$ & $0.766\pm0.066$
& $\rho_{1}$ & $0.522\pm0.006$
& $\rho_{1}$ & $0.437\pm0.030$ \\
& $\rho_{1/2}$ & $0.859\pm0.062$
& $\rho_{1/2}$ & $0.524\pm0.007$
& $\rho_{1/2}$ & $0.547\pm0.026$ \\
& $\widetilde{\rho}_{2}$ & $0.971\pm0.015$
& $\widetilde{\rho}_{1}$ & $\bm{0.9999\pm0.0001}$
& $\widetilde{\rho}_{2}$ & $\bm{0.677\pm0.003}$ \\
& $\widehat{\rho}_{2}$  & $\bm{0.998\pm0.001}$
& $\widehat{\rho}_{1}$  & $0.597\pm0.052$
& $\widehat{\rho}_{2}$ & $0.645\pm0.012$ \\
\arrayrulecolor{gray!80}\midrule[0.5pt]
\multirow{4}{*}{\emph{Algorithm \ref{kmeans_algo}}~~~~} 
& $\rho_{1}$ & $0.758\pm 0.069$
& $\rho_{1}$ & $0.516\pm 0.002$
& $\rho_{1}$  & $0.452\pm 0.030$ \\
& $\rho_{1/2}$ & $0.901\pm 0.060$
& $\rho_{1/2}$ & $0.524\pm 0.007$
& $\rho_{1/2}$ & $0.570\pm 0.016$ \\
& $\widetilde{\rho}_{2}$ & $0.971\pm 0.015$
& $\widetilde{\rho}_{1}$ & $\bm{0.9999\pm 0.0001}$
& $\widetilde{\rho}_{2}$ & $\bm{0.673\pm 0.002}$ \\
& $\widehat{\rho}_{2}$ & $\bm{0.998\pm 0.001}$
& $\widehat{\rho}_{1}$ & $0.528\pm 0.008$
& $\widehat{\rho}_{2}$ & $0.640\pm 0.013$ \\
\arrayrulecolor{gray!80}\midrule[0.5pt]
\emph{$k$-means}~~~~ 
& & $0.599\pm0.046$
& & $0.521\pm0.005$
& & $0.360\pm0.004$ \\
\emph{GMM}~~~~
& & $0.9995\pm0.0003$
& & $0.598\pm0.018$
& & $0.479\pm0.021$ \\
\arrayrulecolor{black}\bottomrule[1pt]
\end{tabular}
\caption{\label{table:other}
Clustering the data shown in Fig.~\ref{fig:other} with Algorithm~\ref{algo}
and Algorithm~\ref{kmeans_algo}, with kernels 
\eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}, as well as $k$-means and GMM. 
We sample $10$ times and show the average
accuracy \eqref{eq:accuracy} with standard error.
}
\end{table}

Next we consider 
the well-known MNIST handwritten digit dataset 
as illustrated in Fig.~\ref{fig:other}d.
Each data point is an $8$-bit gray scale
image forming a $784$-dimensional vector 
corresponding to the digits $\{0,1,\dotsc,9 \}$.
Besides the kernel \eqref{eq:rhoalpha}, we
consider the gaussian kernel \eqref{eq:rhogsigma} with 
\begin{equation}
\label{eq:sigma}
\sigma^2 = \dfrac{1}{n^2} \sum_{i,j=1}^n \| x_i - x_j \|^2 ,
\end{equation}
which is computed from a sample  $\{ x_i \}_{i=1}^n$.
% @gui: it is cheating if this sample is part of the data, this should be a sample of "training data", not used for subsequent testing. please clarify
We consider subsets of the  classes $\{0,1,\dotsc,9 \}$, 
where we sample $100$ points 
for each class. We perform clustering through 
Algorithm~\ref{algo},
Algorithm~\ref{kmeans_algo}, and $k$-means.
The results are shown in Table~\ref{table:mnist} where the kernel
and its parameter for each case is indicated.
Algorithm~\ref{algo} performed slightly better than $k$-means, except
for the last column where all the methods are comparable. 
Unsupervised clustering on MNIST dataset without any feature extraction
or dimensionality reduction is not an easy task. For instance,
the same experiment was performed in \cite{Sapiro} where a low-rank
transformation is learned then subsequently used in subspace clustering,
providing very accurate results. One could explore analogous methods
for learning a better representation of the data and subsequently apply
Algorithm~\ref{algo} for clustering.

\newcolumntype{g}{>{\columncolor{gray!20}}l}
\begin{table}[h]
\renewcommand*{\arraystretch}{0.75}
\begin{tabular}{@{}rl|l|l|l|l@{}}
\toprule[1pt]
\emph{Class Subset}~~~~ & & $\{0,1,2\}$ &
$\{0,1,\dotsc,4\}$ &
$\{0,1,\dotsc,6\}$ &
$\{0,1,\dotsc,8\}$ \\
\midrule[0.5pt]
\multirow{3}{*}{\emph{Algorithm \ref{algo}}~~~~} 
& $\rho_{1}$\hspace{1em} 
&$0.907\pm 0.007$
&$0.866\pm 0.006$
&$0.715\pm 0.013$
&$0.616\pm 0.019$
\\
& $\rho_{1/2}$ 
&$\bm{0.918\pm 0.006}$
&$0.849\pm 0.025$
&$0.711\pm 0.010$
&$0.642\pm 0.009$
\\
& $\widetilde{\rho}_{\sigma}$ 
&$0.900\pm 0.007$
&$\bm{0.871\pm 0.005}$
&$\bm{0.719\pm 0.010}$
&$0.630\pm 0.016$
\\
\arrayrulecolor{gray!80}\midrule[0.5pt]
\multirow{3}{*}{\emph{Algorithm \ref{kmeans_algo}}~~~~} 
& $\rho_{1}$ 
&$0.914\pm 0.006$
&$0.845\pm 0.023$
&$0.664\pm 0.022$
&$0.614\pm 0.014$
\\
& $\rho_{1/2}$ 
&$0.895\pm 0.011$
&$0.822 \pm 0.026$
&$0.669\pm 0.021$
&$0.591\pm 0.019$
\\
& $\widetilde{\rho}_{\sigma}$ 
&$0.896\pm 0.007$
&$0.869\pm 0.006$
&$0.705\pm 0.016$
&$\bm{0.646\pm 0.020}$
\\
\arrayrulecolor{gray!80}\midrule[0.5pt]
\emph{$k$-means}~~~~ &
&$0.871\pm 0.015$
&$0.840\pm 0.022$
&$0.707\pm 0.012$
&$0.634\pm 0.011$
\\
\arrayrulecolor{black}\bottomrule[1pt]
\end{tabular}
% @gui: i'm not convinced you need alg 2 in these tables at all. if i were to add alg 2 to these tables, however, i would do the same thing i would for the figs, that is, for each simulation, compute the diff between performance on alg 2 and alg 3, and report that.
\caption{\label{table:mnist}
Clustering the data shown in Fig.~\ref{fig:other}d with Algorithm~\ref{algo},
Algorithm~\ref{kmeans_algo}, and $k$-means.
We use the kernel \eqref{eq:rhoalpha} with $\alpha\in\{1,2\}$ and
the gaussian kernel 
\eqref{eq:rhogsigma} with $\sigma$ given by \eqref{eq:sigma}. 
For each subset of digits we sample $10$ times and show the average
accuracy \eqref{eq:accuracy} with standard error. We sample $100$ points for 
each class.
}
\end{table}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\label{sec:conclusion}
